\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2025-04-08 misc/JSS article/R&R/article.Rnw
Fix computation of RCA indices.

<<exports>>=
set.seed(101)
exports <- expand.grid(y=1:10, c=paste0("c",1:10), s=paste0("s",1:10)) |>
  tfm(v = abs(rnorm(1e3))) |> colorder(c, s) |> ss(-sample.int(1e3, 500))
@
Like any real trade dataset, it is unbalanced---\code{ss(-sample.int(1e3, 500))} randomly removes 500 rows. Suppose we wanted to extract the latest trade within the last two years.
<<exports_2>>=
latest <- fsubset(exports, y >= 8 & y == fmax(y, list(c, s), "fill"), -y)
@
Below, I compute how many products different countries have exported in the last two years
<<prod>>=
with(latest, fndistinct(s, c))
@
and this computes \citet{balassa1965tariff}'s index of Revealed Comparative Advantage, defined as the share of a sector in country exports divided by the share of the sector in world exports.
<<RCA>>=
with(latest, fsum(v, c, TRA = "/") / fsum(v, s, TRA = "/"))[1:10]
@
More complex use cases are frequent in my work. For example, I recently combined multiple spatial datasets on points of interest (POIs). In the face of significant duplicates and problems matching POIs directly across datasets, I decided to keep the richest source for each location and POI type. After creating POI confidence, location, and type indicators comparable across sources (datasets), my deduplication expression was \code{fsubset(data\_long, source == fmode(source, list(location, type), confidence, "fill"))}---which retains POIs from the confidence-weighted most frequent (i.e., richest) source by location and type.

@@

Last but not least, I benchmark pivots, starting a with long pivot that simply melt the 6 columns aggregated beforehand into one column, duplicating all other columns 6 times:
%
<<bench_pivot_long, cache=TRUE, max.print=50>>=
bmark(tidyr = tidyr::pivot_longer(flights, cols = vars),
      data.table = qDT(flights) |> melt(measure = vars),
      collapse = pivot(flights, values = vars))
@
%
Memory-wise, \pkg{collapse} and \pkg{data.table} are equally efficient, but \pkg{collapse} is faster, presumably due to more extensive use of \code{memset()} to copy values in \proglang{C}, or smaller \proglang{R}-level overheads.

>>>>

<<exports>>=
set.seed(101)
exports <- expand.grid(y=1:10, c=paste0("c",1:10), s=paste0("s",1:10)) |>
  tfm(v = abs(rnorm(1e3))) |> colorder(c, s) |> ss(-sample.int(1e3, 500))
@
Like any real trade dataset, it is unbalanced---\code{ss(-sample.int(1e3, 500))} randomly removes 500 rows. Suppose we wanted to extract the latest trade within the last two years.
<<exports_2>>=
latest <- fsubset(exports, y >= 8 & y == fmax(y, list(c, s), "fill"), -y)
@
Below, I compute how many products different countries have exported in the last two years
<<prod>>=
with(latest, fndistinct(s, c))
@
and this computes \citet{balassa1965tariff}'s index of Revealed Comparative Advantage, defined as the share of a sector in country exports divided by the share of the sector in world exports.
<<RCA>>=
with(latest, fsum(v, c, TRA = "/") / fsum(fsum(v, TRA = "/"), s, TRA = "fill"))[1:10]
@
More complex use cases are frequent in my work. For example, I recently combined multiple spatial datasets on points of interest (POIs). In the face of significant duplicates and problems matching POIs directly across datasets, I decided to keep the richest source for each location and POI type. After creating POI confidence, location, and type indicators comparable across sources (datasets), my deduplication expression was \code{fsubset(data\_long, source == fmode(source, list(location, type), confidence, "fill"))}---which retains POIs from the confidence-weighted most frequent (i.e., richest) source by location and type.

@@

Last but not least, I benchmark pivots, starting a with long pivot that simply melts the 6 columns aggregated beforehand into one column, duplicating all other columns 6 times:
%
<<bench_pivot_long, cache=TRUE, max.print=50>>=
bmark(tidyr = tidyr::pivot_longer(flights, cols = vars),
      data.table = qDT(flights) |> melt(measure = vars),
      collapse = pivot(flights, values = vars))
@
%
Memory-wise, \pkg{collapse} and \pkg{data.table} are equally efficient, but \pkg{collapse} is faster, presumably due to more extensive use of \code{memset()} to copy values in \proglang{C}, or smaller \proglang{R}-level overheads.
